---
title: "*Spatial Statistics Lab 2*"
author: '**Onyedikachi J Okeke**'
date: "`r Sys.Date()`"
output:
  word_document: default
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## **0.0 Load and summarize the data set**

```{r,comment="",echo=TRUE,eval=FALSE}
# Specify the file path
file_path <- "C:/Users/GIS/Desktop/UNM Doc/STATISTICS PROGRAM/MY COURSES/Spring 2024 Courses/Spatial Statistics/Lab work/Lab 2/Lab2data.csv"

# Load the data into R as Lab2data
Lab2data <- read.csv(file_path)

# Print the loaded data
summary(Lab2data)
```
#### Result
```{r,echo=F,results='Show', fig.show='hold',eval=T}
# Specify the file path
file_path <- "C:/Users/GIS/Desktop/UNM Doc/STATISTICS PROGRAM/MY COURSES/Spring 2024 Courses/Spatial Statistics/Lab work/Lab 2/Lab2data.csv"

# Load the data into R as Lab2data
Lab2data <- read.csv(file_path)

# Print the loaded data
summary(Lab2data)
```

## **1.0 Find the dimensions of the loaded data**

```{r,comment="",echo=TRUE,eval=FALSE}
# Print the dimensions of the loaded data
cat("Dimensions of Lab2data:", dim(Lab2data), "\n")
```
#### Result
```{r,echo=F,results='Show', fig.show='hold',eval=T}

# Print the dimensions of the loaded data
cat("Dimensions of Lab2data:", dim(Lab2data), "\n")

```

## **2.0 What are the pollen count values for the first 10 days?**

```{r,comment="",echo=TRUE,eval=FALSE}
# Count values for the first 10 days
pollen_count <- Lab2data$pollen[1:10]
print(pollen_count)

```
#### Result
```{r,echo=F,results='Show', fig.show='hold',eval=T}

# Count values for the first 10 days
pollen_count <- Lab2data$pollen[1:10]
print(pollen_count)

```

## **3.0 Calculate the following statistics for pollen count**

```{r,comment="",echo=TRUE,eval=FALSE}
# Assuming "pollen_count" is the column name in Lab2data
pollen_count <- Lab2data$pollen

# Calculate mean
mean_pollen <- mean(pollen_count)

# Calculate variance
var_pollen <- var(pollen_count)

# Calculate standard deviation
std_dev_pollen <- sd(pollen_count)

# Calculate quantiles (e.g., 25th, 50th, and 75th percentiles)
quantiles_pollen <- quantile(pollen_count, c(0.25, 0.50, 0.75))

# Print the results
cat("Mean Pollen Count:", mean_pollen, "\n")
cat("Variance of Pollen Count:", var_pollen, "\n")
cat("Standard Deviation of Pollen Count:", std_dev_pollen, "\n")
cat("Quantiles of Pollen Count (25%, 50%, 75%):", quantiles_pollen, "\n")
```
#### Result
```{r,echo=F,results='Show', fig.show='hold',eval=T}

# Assuming "pollen_count" is the column name in Lab2data
pollen_count <- Lab2data$pollen

# Calculate mean
mean_pollen <- mean(pollen_count)

# Calculate variance
var_pollen <- var(pollen_count)

# Calculate standard deviation
std_dev_pollen <- sd(pollen_count)

# Calculate quantiles (e.g., 25th, 50th, and 75th percentiles)
quantiles_pollen <- quantile(pollen_count, c(0.25, 0.50, 0.75))

# Print the results
cat("Mean Pollen Count:", mean_pollen, "\n")
cat("Variance of Pollen Count:", var_pollen, "\n")
cat("Standard Deviation of Pollen Count:", std_dev_pollen, "\n")
cat("Quantiles of Pollen Count (25%, 50%, 75%):", quantiles_pollen, "\n")

```

## **4.0 Statistics for pollen count in April?**

```{r,comment="",echo=TRUE,eval=FALSE}
# Assuming "Date" is the column name in Lab2data for dates, and "pollen_count" for pollen counts

# Filter data for April
april_data <- subset(Lab2data,Month == "April")

# Assuming "pollen_count" is the column name for pollen counts
pollen_count_april <- april_data$pollen

#print data
print(pollen_count_april)

# Calculate mean
mean_pollen_april <- mean(pollen_count_april)

# Calculate variance
var_pollen_april <- var(pollen_count_april)

# Calculate standard deviation
std_dev_pollen_april <- sd(pollen_count_april)

# Calculate quantiles (e.g., 25th, 50th, and 75th percentiles)
quantiles_pollen_april <- quantile(pollen_count_april, c(0.25, 0.50, 0.75))

# Print the results
cat("Mean Pollen Count in April:", mean_pollen_april, "\n")
cat("Variance of Pollen Count in April:", var_pollen_april, "\n")
cat("Standard Deviation of Pollen Count in April:", std_dev_pollen_april, "\n")
cat("Quantiles of Pollen Count in April (25%, 50%, 75%):", quantiles_pollen_april, "\n")

```
#### Result
```{r,echo=F,results='Show', fig.show='hold',eval=T}

# Assuming "Date" is the column name in Lab2data for dates, and "pollen_count" for pollen counts

# Filter data for April
april_data <- subset(Lab2data,Month == "April")

# Assuming "pollen_count" is the column name for pollen counts
pollen_count_april <- april_data$pollen

#print data
print(pollen_count_april)

# Calculate mean
mean_pollen_april <- mean(pollen_count_april)

# Calculate variance
var_pollen_april <- var(pollen_count_april)

# Calculate standard deviation
std_dev_pollen_april <- sd(pollen_count_april)

# Calculate quantiles (e.g., 25th, 50th, and 75th percentiles)
quantiles_pollen_april <- quantile(pollen_count_april, c(0.25, 0.50, 0.75))

# Print the results
cat("Mean Pollen Count in April:", mean_pollen_april, "\n")
cat("Variance of Pollen Count in April:", var_pollen_april, "\n")
cat("Standard Deviation of Pollen Count in April:", std_dev_pollen_april, "\n")
cat("Quantiles of Pollen Count in April (25%, 50%, 75%):", quantiles_pollen_april, "\n")

```

## **5.0 Histograms (in frequencies) of pollen count in April, May, and June**

```{r,comment="",echo=TRUE,eval=FALSE}
# Assuming "Month" and "pollen_count" are the column names in Lab2data
# Assuming "Date" is already in Date format

# Subset data for April, May, and June
may_data <- subset(Lab2data,Month == "May")
pollen_count_may <- may_data$pollen

june_data <- subset(Lab2data,Month == "June")
pollen_count_june <- june_data$pollen

# Create a 1x3 layout for the histograms
par(mfrow = c(1, 3))

# Plot histogram for April
hist(pollen_count_april, main = "Pollen Count in April", xlab = "Pollen Count", ylab = "Frequency")

# Plot histogram for May
hist(pollen_count_may, main = "Pollen Count in May", xlab = "Pollen Count", ylab = "Frequency")

# Plot histogram for June
hist(pollen_count_june, main = "Pollen Count in June", xlab = "Pollen Count", ylab = "Frequency")

# Reset the layout to default
par(mfrow = c(1, 1))

```
#### Result
```{r,echo=F,results='Show', fig.show='hold',eval=T}

# Assuming "Month" and "pollen_count" are the column names in Lab2data
# Assuming "Date" is already in Date format

# Subset data for April, May, and June
may_data <- subset(Lab2data,Month == "May")
pollen_count_may <- may_data$pollen

june_data <- subset(Lab2data,Month == "June")
pollen_count_june <- june_data$pollen

# Create a 1x3 layout for the histograms
par(mfrow = c(1, 3))

# Plot histogram for April
hist(pollen_count_april, main = "Pollen Count in April", xlab = "Pollen Count", ylab = "Frequency")

# Plot histogram for May
hist(pollen_count_may, main = "Pollen Count in May", xlab = "Pollen Count", ylab = "Frequency")

# Plot histogram for June
hist(pollen_count_june, main = "Pollen Count in June", xlab = "Pollen Count", ylab = "Frequency")

# Reset the layout to default
par(mfrow = c(1, 1))

```

## **6.0 Use qqnorm and qqline to construct plots for temperature and pollen count**

```{r,comment="",echo=TRUE,eval=FALSE}
par(mfrow=c(1,2))

# QQ plot for temperature
qqnorm(Lab2data$Temp2, main = "QQ Plot for Temperature")
qqline(Lab2data$Temp2, col = 2)


# QQ plot for pollen count
qqnorm(Lab2data$pollen, main = "QQ Plot for Pollen Count")
qqline(Lab2data$pollen, col = 2)


# Shapiro-Wilk Normality Test for temperature
shapiro_temp <- shapiro.test(Lab2data$Temp2)
cat("Shapiro-Wilk Normality Test for Temperature:\n", "p-value =", shapiro_temp$p.value, "\n")

# Shapiro-Wilk Normality Test for pollen count
shapiro_pollen <- shapiro.test(Lab2data$pollen)
cat("Shapiro-Wilk Normality Test for Pollen Count:\n", "p-value =", shapiro_pollen$p.value, "\n"

```
#### Result
```{r,echo=F,results='Show', fig.show='hold',eval=T}

par(mfrow=c(1,2))

# QQ plot for temperature
qqnorm(Lab2data$Temp2, main = "QQ Plot for Temperature")
qqline(Lab2data$Temp2, col = 2)


# QQ plot for pollen count
qqnorm(Lab2data$pollen, main = "QQ Plot for Pollen Count")
qqline(Lab2data$pollen, col = 2)


# Shapiro-Wilk Normality Test for temperature
shapiro_temp <- shapiro.test(Lab2data$Temp2)
cat("Shapiro-Wilk Normality Test for Temperature:\n", "p-value =", shapiro_temp$p.value, "\n")

# Shapiro-Wilk Normality Test for pollen count
shapiro_pollen <- shapiro.test(Lab2data$pollen)
cat("Shapiro-Wilk Normality Test for Pollen Count:\n", "p-value =", shapiro_pollen$p.value, "\n")

```

The Shapiro-Wilk Normality Test is used to assess whether a sample comes from a normally distributed population. In this case, the p-values obtained for both the temperature and pollen count are extremely small:

Temperature:

Shapiro-Wilk p-value: 0.0002311045
Pollen Count:

Shapiro-Wilk p-value: 2.086788e-23 (very close to zero)

The small p-values in both cases suggest that we have evidence to reject the null hypothesis that the data follows a normal distribution. In other words, the data for both temperature and pollen count are likely not normally distributed.


## **7.0 Plot to construct two graphs for pollen count and EVI for the whole dataset**

```{r,comment="",echo=TRUE,eval=FALSE}

# Convert the "Day" column to a Date type
Lab2data$Day <- as.Date(Lab2data$Day, format="%m/%d/%Y")

# Plot 1: Pollen Count
plot(Lab2data$Day, Lab2data$pollen, type="l", col="blue", xlab="Date", ylab="Pollen Count", main="Pollen Count Over Time")

# Plot 2: EVI
plot(Lab2data$Day, Lab2data$EVI, type="l", col="green", xlab="Date", ylab="EVI", main="EVI Over Time")

```
#### Result
```{r,echo=F,results='Show', fig.show='hold',eval=T}

# Convert the "Day" column to a Date type
Lab2data$Day <- as.Date(Lab2data$Day, format="%m/%d/%Y")


# Plot 1: Pollen Count
plot(Lab2data$Day, Lab2data$pollen, type="l", col="blue", xlab="Date", ylab="Pollen Count", main="Pollen Count Over Time")

# Plot 2: EVI
plot(Lab2data$Day, Lab2data$EVI, type="l", col="green", xlab="Date", ylab="EVI", main="EVI Over Time")

```

## **8.0 two-sample t-test to test**

```{r,comment="",echo=TRUE,eval=FALSE}

# Perform a two-sample t-test
t_test_result <- t.test(pollen_count_april, pollen_count_may)
print(t_test_result)

```
#### Result
```{r,echo=F,results='Show', fig.show='hold',eval=T}
# Perform a two-sample t-test
t_test_result <- t.test(pollen_count_april, pollen_count_may)
print(t_test_result)

```

## **9.0 two-sample t-test to test**

```{r,comment="",echo=TRUE,eval=FALSE}

# Fit a linear regression model
regression_model <- lm(pollen ~ EVI, data = Lab2data)

# Get the formula and coefficients
formula <- as.formula(regression_model)
coefficients <- coef(regression_model)

# Print the formula and coefficients
cat("Regression Formula:", format(formula), "\n")
cat("Coefficients:\n")
print(coefficients)

# Check the significance of the predictor (EVI)
summary(regression_model)

par(mfrow=c(1,2))

# Plot the data and add the fitted regression line
plot(Lab2data$EVI, Lab2data$Pollen, xlab = "EVI", ylab = "Pollen Count", main = "Pollen Count Vs EVI")
abline(regression_model, col = "red")

# Plot residuals in a separate graph
residuals <- residuals(regression_model)
plot(Lab2data$EVI, residuals, xlab = "EVI", ylab = "Residuals", main = "Residuals Vs EVI
```
#### Result
```{r,echo=F,results='Show', fig.show='hold',eval=T}

# Fit a linear regression model
regression_model <- lm(pollen ~ EVI, data = Lab2data)

# Get the formula and coefficients
formula <- as.formula(regression_model)
coefficients <- coef(regression_model)

# Print the formula and coefficients
cat("Regression Formula:", format(formula), "\n")
cat("Coefficients:\n")
print(coefficients)

# Check the significance of the predictor (EVI)
summary(regression_model)

par(mfrow=c(1,2))

# Plot the data and add the fitted regression line
plot(Lab2data$EVI, Lab2data$Pollen, xlab = "EVI", ylab = "Pollen Count", main = "Pollen Count Vs EVI")
abline(regression_model, col = "red")

# Plot residuals in a separate graph
residuals <- residuals(regression_model)
plot(Lab2data$EVI, residuals, xlab = "EVI", ylab = "Residuals", main = "Residuals Vs EVI")

```

## **10.0 two-sample t-test to test**

```{r,comment="",echo=TRUE,eval=FALSE}
# Assuming "Pollen," "EVI," "Precipitation," and "Temperature" are the correct column names in Lab2data
# Fit a multiple linear regression model
regression_model <- lm(pollen ~ EVI + preci + Temp2, data = Lab2data)

# Get the formula and coefficients
formula <- as.formula(regression_model)
coefficients <- coef(regression_model)

# Print the formula and coefficients
cat("Regression Formula:", format(formula), "\n")
cat("Coefficients:\n")
print(coefficients)

# Check the significance of the predictors
summary(regression_model)

# Plot residuals
residuals <- residuals(regression_model)
plot(residuals, xlab = "Observation", ylab = "Residuals", main = "Residuals Plot")
abline(h = 0, col = "red", lty = 2)  # Add a horizontal line at y = 0

```
#### Result
```{r,echo=F,results='Show', fig.show='hold',eval=T}
# Assuming "Pollen," "EVI," "Precipitation," and "Temperature" are the correct column names in Lab2data
# Fit a multiple linear regression model
regression_model <- lm(pollen ~ EVI + preci + Temp2, data = Lab2data)

# Get the formula and coefficients
formula <- as.formula(regression_model)
coefficients <- coef(regression_model)

# Print the formula and coefficients
cat("Regression Formula:", format(formula), "\n")
cat("Coefficients:\n")
print(coefficients)

# Check the significance of the predictors
summary(regression_model)

# Plot residuals
residuals <- residuals(regression_model)
plot(residuals, xlab = "Observation", ylab = "Residuals", main = "Residuals Plot")
abline(h = 0, col = "red", lty = 2)  # Add a horizontal line at y = 0

```